---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

- Load the data
```{r}
d <- read.csv(file="activity.csv", header=TRUE, sep=",", stringsAsFactors = FALSE, na.strings = 'NA')
```
- Convert data column to Date type:
```{r}
d$date <-  as.Date(d$date, format = "%Y-%m-%d")
str(d)
```

## What is mean total number of steps taken per day?
- Calculate total number of steps taken each day
```{r}
spd <- tapply(d$steps, d$date, sum, na.rm = TRUE)
head(spd)
```

- Plot histogram of the total number of steps taken each day
```{r}
hist(spd,  main = "Total # of steps taken per day", xlab = "# of steps")
```

- Calculate mean and median total number of steps taken per day
```{r}
mean(spd)
median(spd)
```
## What is the average daily activity pattern?
- Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
sp5m <- tapply(d$steps, d$interval, mean, na.rm = TRUE)
plot(names(sp5m), sp5m, type="l", ylab = "# of steps", xlab = "Interval")
```

- Use which.max function to get the 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps
```{r}
names(which.max(sp5m))
```

## Imputing missing values
Since there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

- Total number of missing values in the dataset 
```{r}
sum(is.na(d$steps))
```

- We fill in all of the missing values with the mean for that 5-minute interval (from sp5m: we have calculated above)
```{r}
for (i in 1:nrow(d)){ if (is.na(d[i,1])){ d[i,1] <- sp5m[as.character(d[i,3])]; } }
```

- Now all missing values are filled.
```{r}
head(d)
```

- Re-calculate total number of steps taken each day and plot new histogram 
```{r}
spd2 <- tapply(d$steps, d$date, sum)
hist(spd2,  main = "Total # of steps taken per day (removed NAs)", xlab = "# of steps")
```

- Re-calculate mean and median
```{r}
mean(spd2)
median(spd2)
```

## Are there differences in activity patterns between weekdays and weekends?

- Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
day_type <- ifelse(weekdays(d$date) %in% c("Sunday","Saturday") ,"weekend", "weekday")
head(day_type)
```

- Calculate 5-minute interval with day_type as factor
```{r}
sp5m2 <- setNames(aggregate(d$step ~ d$interval + day_type, data = d, mean), c("Interval", "DayType", "Steps"))
sp5m2 <- transform(sp5m2, DayType = factor(DayType))
```

- Show a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r}
library(lattice)
xyplot(data = sp5m2, Steps ~ Interval | DayType, layout = c(1,2), type = "l", ylab = "Number of steps")
```